Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-02-29 article_springer/article_springer.tex
Cleaned Up

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)
% Pandoc header

@@

  \title{Statistical Inference for inter-arrival times of extreme events in
bursty time series}
    \author[a]{Katharina Hees}
   \ead{hees@statistik.tu-dortmund.de}

    \author[b]{Smarak Nayak}
   \ead{smarak.nayak@nab.com.au}

    \author[c]{Peter Straka}
   \ead{p.straka@unsw.edu.au}

      \address[a]{Department of Statistics, TU Dortmund University, Dortmund, Germany}
    \address[b]{National Australia Bank, Melbourne, Australia}
    \address[c]{School of Mathematics and Statistics, UNSW, Sydney, Australia}

@@

Time series displaying temporally inhomogeneous behaviour in terms of
the occurrence of events have received strong interest in the recent
statistical physics literature (Barabási 2005; Oliveira and Barabási
2005; Vasquez et al. 2006; Vazquez et al. 2007; Omi and Shinomoto 2011;
Min, Goh, and Vazquez 2011; Karsai et al. 2011; Bagrow and Brockmann
2013). They have been observed in the context of earthquakes, sunspots,
neuronal activity and human communication (see Karsai et al. 2012;
Vajna, Tóth, and Kertész 2013; Meerschaert and Stoev 2008 for a list of
references). Such time series exhibit high activity in some `bursty'
intervals, which alternate with other, quiet intervals. Although several
mechanisms are plausible explanations for bursty behaviour (most
prominently self-exciting point process by Hawkes (1971)), there seems
to be one salient feature which very typically indicates the departure
from temporal homogeneity: a heavy-tailed distribution of waiting times
(Vasquez et al. 2006; Karsai et al. 2012; Vajna, Tóth, and Kertész
2013). As we show below in simulations, a simple renewal process with
heavy-tailed waiting times can capture this type of dynamics. For many
systems, the renewal property is appropriate; a simple test of the
absence of correlations in a succession of waiting times can be
undertaken by randomly reshuffling the waiting times (Karsai et al.
2012).

Often a magnitude, or mark can be assigned to each event in the renewal
process, such as for earthquakes, solar flares or neuron voltages. The
Peaks-Over-Threshold model (POT, see e.g.~Coles 2001) applies a
threshold to the magnitudes, and fits a Generalized Pareto distribution
to the threshold exceedances. A commonly made assumption in POT models
is that times between events are either fixed or light-tailed, and this
entails that the threshold crossing times form a Poisson process (``On
the Exceedance Point Process for a Stationary Sequence'' 1988). Then as
one increases the threshold and thus decreases the threshold crossing
probability \(p\), the Poisson process is rarefied, i.e.~its intensity
decreases \emph{linearly} with \(p\) (see e.g.~Beirlant et al. 2006).

As will be shown below, in the heavy-tailed waiting time scenario
threshold crossing times form a \emph{fractional Poisson process}
(Laskin 2003; Meerschaert, Nane, and Vellaisamy 2011), which is a
renewal process with Mittag-Leffler distributed waiting times. The
family of Mittag-Leffler distributions nests the exponential
distribution (Haubold, Mathai, and Saxena 2011), and hence the
fractional Poisson process generalizes the standard Poisson process.
Again as the threshold size increases and the threshold crossing
probability \(p\) decreases, the fractional Poisson process is rarefied:
The scale parameter of the Mittag-Leffler inter-arrival times of
threshold crossing times increases, but \emph{superlinearly}; see the
Theorem below.

Maxima of events which occur according to a renewal process with
heavy-tailed waiting times have been studied under the names
``Continuous Time Random Maxima process'' (CTRM) (Benson, Schumer, and
Meerschaert 2007; Meerschaert and Stoev 2008; Hees and Scheffler 2018b,
2018a), ``Max-Renewal process'' (Silvestrov 2002; Silvestrov and Teugels
2004; Basrak and Špoljarić 2015), and ``Shock process'' (Esary and
Marshall 1973; Shanthikumar and Sumita 1983, 1984, 1985; Anderson 1987;
Gut and Hüsler 1999). The existing literature focuses on probabilistic
results surrounding these models. In this work, however, we introduce a
method of inference for this type of model, which is seemingly not
available in the literature.

We review the marked renewal process in Section 2, and derive a scaling
limit theorem for inter-exceedance times in Section 3. We give a
statistical procedure to estimate model parameters via stability plots
in Section 5, but to set the stage we first discuss inference for the
Mittag-Leffler distribution in Section 4. A simulation study of the
effectiveness of our statistical procedure is given in Section 6. In
Section 7 we apply our method to a real data set. In Section 8, we
discuss the memory property of the Mittag-Leffler distribution, and how
it affects the predictive distribution for the time until the next
threshold crossing event. Finally we close with a discussion and
conclusion in Section 9. For all statistical computations we have used R
(R Core Team 2018). All code and data used for the analysis in this
article has been organized into an R package \texttt{CTRE}
(\url{https://github.com/strakaps/CTRE}). The source code for the
figures generated in this manuscript is available online at
\url{https://github.com/strakaps/bursty-POT}.

@@

In this article, we restrict ourselves to the uncoupled case, where
\(W\) and \(J\) are independent. Then the two sequences
\(X(\ell, n)_{n \in \mathbb N}\) and \(T(\ell, n)_{n \in \mathbb N}\)
are independent as well. To see why, note that \(X(\ell)\) is, in
distribution, simply equal to \(J | J > \ell\), independent of any
waiting time \(W_k\).\\
Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top
panel, where \(W\) has a stable distribution with tail parameter
\(\beta =\) 0.8 (and skewness \(1\) and location \(0\)), and where \(J\)
is from a standard Gumbel distribution. In the bottom panel, we plot a
time series of solar flare intensities derived from a NASA dataset
(Dennis et al. 1991) which we will later examine more closely (see
Section 7). Clearly, the simulated data exhibit long intervals
\emph{without any} events, whereas in the real-world dataset events
appear continuously. The threshold exceedances, however, appear to have
visually similar statistical behaviour in both models. Observations
below a threshold are commonly discarded in Extreme Value Theory (POT
approach); likewise, the CTRE model interprets these observations as
noise and discards them.

@@

In this section we state and prove the key theorem, see below. For an
accessible introduction to regular variation and stable limit theorems,
we recommend the book by Mark M Meerschaert and Sikorskii (2012).

@@

\emph{Proof of Theorem:} We interpret the threshold crossing time
\(T(\ell)\) as the hitting time of the underlying CTRM (Continuous Time
Random Maxima) or ``max-renewal process'', and then utilize a result by
Meerschaert and Stoev (2008). The running maximum process is defined as
\[
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
\] and since we assume that the \(J_k\) have a continuous distribution,
there exist norming functions \(a(c)\) and \(d(c)\) such that \[
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
\] where \(F\) is a generalized extreme value distribution, and
\(\ell^*\) is any value from the support of \(F\). The CTRM process is
then defined via \[
V(t) = M(N(t)), \quad t \ge 0
\] where \(N(t)\) is the renewal process associated with the waiting
times \(W_k\): \[
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
\] Now a key observation is that \[
T(\ell) = \inf\{t: V(t) > \ell\}, 
\] and that \[
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
\] By (Theorem 3.1, Meerschaert and Stoev 2008), we have the stochastic
process convergence \[
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
\] where \(Y(t)\) is a time-changed (``subordinated'') extremal process,
and where \(\tilde b(c)\) is a regularly varying norming function which
is \emph{inverse} to \(b(c)\), in the sense that
\(b(\tilde b(c)) \sim c \sim \tilde b(b(c))\).

Without loss of generality, we choose \(\ell^*\) such that
\(F(\ell^*) = 1/e\), and let
\(\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))\). We may then calculate
\[
\mathbf P\left[ \frac{T(\ell)}{b(1/p)} > t \right]
= \mathbf P[T(\ell) > b(1/p) t]
= \mathbf P[V(ct) \le \ell]
\] where we have substituted \(c = b(1/p)\). Moreover \[
\mathbf P[V(ct) \le \ell]
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
\] Defining the hitting time of level \(\ell^*\) by \(Y(t)\) as
\(\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}\), we then have \[
P\left[ Y(t) \le \ell^* \right] = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
\] by (Proposition 4.2, Meerschaert and Stoev 2008), where \(X\) is an
exponential random variable with mean \(1\). Using (Theorem 19.1,
Haubold, Mathai, and Saxena 2011), we see that
\(X^{1/\beta} D \sim {\rm ML}(\beta, 1)\), concluding the proof. \qed

For a scale parameter \(\sigma > 0\), we write
\({\rm ML}(\beta, \sigma)\) for the distribution of \(\sigma W_\beta\).
The Mittag-Leffler distribution with parameter \(\beta \in (0,1]\) is a
heavy-tailed positive distribution for \(\beta < 1\), with infinite
mean. However, as \(\beta \uparrow 1\), \({\rm ML}(\beta, \sigma)\)
converges weakly to the exponential distribution \({\rm Exp}(\sigma)\)
with mean \(\sigma\). This means that although its moments are all
infinite, the Mittag-Leffler distribution may (if \(\beta\) is close to
1) be indistinguishable from the exponential distribution, for the
purposes of applied statistics. For a detailed reference on the
Mittag-Leffler distribution, see e.g.~ Haubold, Mathai, and Saxena
(2011), and for algorithms, see e.g.~the R package
\texttt{MittagLeffleR} (Gill and Straka 2017).

@@

Historically, the first method proposed for the estimation of the
Mittag-Leffler distribution parameters was the fractional moment
estimator by Kozubowski (2001). Unlike the first moments, the fractional
moments of order \(p\) for \(p<\beta\) exist and are tractable. One
drawback of this method is that constant priors for the tail parameter
are needed for the calculation of the estimates. Cahoy, Uchaikin, and
Woyczynski (2010) proposed a moment estimator of the log-transformed
data, which does not require any prior. Furthermore, they performed
simulation studies illustrating that the log-moment outperforms the
fractional moment estimator with respect to bias and root mean squared
error (RMSE).

Due to the form of the density function of a Mittag-Leffler distribution
there exists no closed form for the Maximum Likelihood estimator (MLE).
In the R Package \texttt{MittagLeffleR} (Gill and Straka 2017), Maximum
Likelihood estimation is implemented via numerical optimization. The MLE
slightly outperforms the log-moment estimator regarding bias and RMSE
for big enough sample sizes, but is extremely computational intensive.
Figure \ref{fig:MSE} shows that both estimators perform well, even for
small sample sizes.

@@

Since the Mittag-Leffler distribution is heavy-tailed, many researchers
would intuitively give the highest importance to the tail behaviour of
the distribution. Of course, one can also use established tail exponent
estimators for the estimation of the parameter \(\beta\) such as the
Hill estimator (Hill 1975) based on the \(r+1\) upper order statistics
\begin{align}\label{eq:hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align} where \(X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}\)
denote the order statistics in decreasing order of a sequence
\(X_1,\ldots,X_n\). However, these methods are statistically less
efficient since they only use a portion of the information contained in
the data, as mentioned by Kozubowski (2001). Moreover, the hill
estimator requires a tuning parameter, denoted with \(r\) in
\eqref{eq:hill}. Additionaly, the estimator only performs reliably well
for distributions close to Pareto (see Resnick 1997). Figure
\ref{fig:Hillplots} shows Hill plots for Mittag-Leffler simulated data,
with varying sample sizes and tails. To deduce the correct tail
parameter estimates from these plots is virtually impossible. To
estimate the tail parameter of a Mittag-Leffler distribution with the
hill estimator, the sample size has to be much larger and one has to
choose a large tuning parameter \(r\). In case that only events above a
high threshold are recorded, even \(n=1000\) events are unrealistic.
Furthermore the Hill estimator of course completely fails if the
inter-arrival times are exponential distributed and not heavy-tailed. We
will comeback to the problems with the hill estimtator in Section
\ref{Simulationstudy}.

@@

Since the exponential distribution is nested in the Mittag-Leffler
family of distributions, a Likelihood-ratio Test (LRT) seems to be an
appropriate way to choose between a model with exponential and
Mittag-Leffler inter-exceedance times. Although the two models are
nested, the asymptotic distribution is not \(\chi^2_1\)-distributed, and
Wilk's Theorem does not hold: under \(H_0\), the parameter \(\beta\) of
Mittag-Leffler distribution is equal to \(1\), and hence lies on the
boundary of the parameter space \((0,1]\). Instead, a valid approach is
a bootstrapped Likelihood-ratio test (see e.g.~Davison, Hinkley, and
others 1997). Figure \ref{Fig:LRT} displays the (simulated) power for
the bootstrapped LRT for Mittag-Leffler distributions with varying tail
parameters based on 1000 simulation runs. As expected, the power
decreases for tail parameters close to one, since the Mittag-Leffler
distribution converges as \(\beta \uparrow 1\) to an exponential
distribution; it becomes hard to differentiate a Mittag-Lefller
distribution from an exponential.

@@

The estimates \(\hat \beta\) and \(\hat \sigma_0\) give an estimate of
the distribution of exceedance times, dependent on the threshold
\(\ell\): \begin{align*}
T(\ell) \sim {\rm ML}(\hat \beta, k^{-1/\hat \beta} \hat \sigma_0),
\end{align*} where \(\ell\) equals the \(k\)-th order statistic. For
quick estimates of the Mittag-Leffler parameters we have used the method
of log-transformed moments (Cahoy 2013). We have verified the validity
of our estimation algorithm via simulations, see Section
\ref{Simulationstudy}.

@@

Figure \ref{fig:flares} shows the stability plots for the solar flare
data, on the left for the tail parameter and on the right for the scale
parameter. Dotted lines show 95\% confidence intervals, which are
derived from the asymptotic normality of the log-moments estimators
(Cahoy 2013) and the \(\delta\)-method (Gill and Straka 2017), dashed
lines show the deduced values of \(\beta\) resp.~\(\sigma_0\). The
stability plot for the tail stabilizes nicely round about 0.85 (dashed
line), while it is a little bit harder to deduce an estimate for the
scale parameter. Howerever, the stability plot for the scale parameter
seems to stabilize round about \(3 \times 10^7\) (dashed line).

@@

The bursty CTRE approach to model ``non-Poissonian'' threshold crossing
times should be contrasted with the (now standard) approach of clusters
of extremes, see e.g.~Ferro and Segers (2003). In this approach, i.i.d.
event sequences of magnitudes are generalized to stationary sequences of
event magnitudes (subject to a mixing condition). The two approaches are
fundamentally different: A clustering model assumes that each event
belongs to one particular (latent) group of events. For bursts, however,
the aim is to identify an underlying scale-free pattern in the event
dynamics, which is often characteristic of complex systems. It is an
interesting open problem to develop quality criteria, based e.g.~on
measures of surprise (Lee, Fan, and Sisson 2015), which guide an applied
statistician in the choice between a clustering and a CTRE approach for
a particular problem. Moreover, we believe it may be possible to unify
the two approaches by considering CTREs based on MRPs with a
\emph{stationary}, rather than i.i.d., sequence of magnitudes.

Finally, a purely scale-free pattern for event times may be too rigid as
an assumption for some bursty time series, because often the
heavy-tailed character of the inter-arrival time distribution does not
hold at all time scales; rather, it applies at short and intermediate
time scales, and is truncated (or tempered, reverting to an exponential
distribution) at very long time scales (see e.g.~Mark M. Meerschaert,
Roy, and Shao 2012; and Aban, Meerschaert, and Panorska 2006). In such
situations, a ``tempered'' Mittag-Leffler distribution may provide a
more realistic fit, which we aim to introduce in follow-up work.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Aban06}{}%
Aban, Inmaculada B, Mark M Meerschaert, and Anna K Panorska. 2006.
``Parameter Estimation for the Truncated Pareto Distribution.'' \emph{J.
Am. Stat. Assoc.} 101 (473): 270--77.
\url{https://doi.org/10.1198/016214505000000411}.

\leavevmode\hypertarget{ref-Anderson1987}{}%
Anderson, Kevin K. 1987. ``Limit Theorems for General Shock Models with
Infinite Mean Intershock Times.'' \emph{J. Appl. Probab.} 24 (2):
449--56. \url{http://www.jstor.org/stable/3214268}.

\leavevmode\hypertarget{ref-Bagrow2013}{}%
Bagrow, James P., and Dirk Brockmann. 2013. ``Natural emergence of
clusters and bursts in network evolution.'' \emph{Phys. Rev. X} 3 (2):
1--6. \url{https://doi.org/10.1103/PhysRevX.3.021016}.

\leavevmode\hypertarget{ref-Barabasi2005}{}%
Barabási, Albert László. 2005. ``The origin of bursts and heavy tails in
human dynamics.'' \emph{Nature} 435 (May): 207--11.
\url{https://doi.org/10.1038/nature03459}.

>>>>

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc header

@@

  \title{Statistical Inference for inter-arrival times of extreme events in
bursty time series}
    \author[a]{Katharina Hees}
   \ead{hees@statistik.tu-dortmund.de} 
    \author[b]{Smarak Nayak}
   \ead{smarak.nayak@nab.com.au} 
    \author[c]{Peter Straka}
   \ead{p.straka@unsw.edu.au} 
      \address[a]{Department of Statistics, TU Dortmund University, Dortmund, Germany}
    \address[b]{National Australia Bank, Melbourne, Australia}
    \address[c]{School of Mathematics and Statistics, UNSW, Sydney, Australia}

@@

Time series displaying temporally inhomogeneous behaviour in terms of
the occurrence of events have received strong interest in the recent
statistical physics literature (Bagrow and Brockmann, 2013; Barabási,
2005; Karsai et al., 2011; Min et al., 2011; Oliveira and Barabási,
2005; Omi and Shinomoto, 2011; Vasquez et al., 2006; Vazquez et al.,
2007). They have been observed in the context of earthquakes, sunspots,
neuronal activity and human communication (see Karsai et al., 2012;
Meerschaert and Stoev, 2008 for a list of references; Vajna et al.,
2013). Such time series exhibit high activity in some `bursty'
intervals, which alternate with other, quiet intervals. Although several
mechanisms are plausible explanations for bursty behaviour (most
prominently self-exciting point process by Hawkes (1971)), there seems
to be one salient feature which very typically indicates the departure
from temporal homogeneity: a heavy-tailed distribution of waiting times
(Karsai et al., 2012; Vajna et al., 2013; Vasquez et al., 2006). As we
show below in simulations, a simple renewal process with heavy-tailed
waiting times can capture this type of dynamics. For many systems, the
renewal property is appropriate; a simple test of the absence of
correlations in a succession of waiting times can be undertaken by
randomly reshuffling the waiting times (Karsai et al., 2012).

Often a magnitude, or mark can be assigned to each event in the renewal
process, such as for earthquakes, solar flares or neuron voltages. The
Peaks-Over-Threshold model (POT, see e.g.~Coles, 2001) applies a
threshold to the magnitudes, and fits a Generalized Pareto distribution
to the threshold exceedances. A commonly made assumption in POT models
is that times between events are either fixed or light-tailed, and this
entails that the threshold crossing times form a Poisson process (``On
the Exceedance Point Process for a Stationary Sequence,'' 1988). Then as
one increases the threshold and thus decreases the threshold crossing
probability \(p\), the Poisson process is rarefied, i.e.~its intensity
decreases \emph{linearly} with \(p\) (see e.g.~Beirlant et al., 2006).

As will be shown below, in the heavy-tailed waiting time scenario
threshold crossing times form a \emph{fractional Poisson process}
(Laskin, 2003; Meerschaert et al., 2011), which is a renewal process
with Mittag-Leffler distributed waiting times. The family of
Mittag-Leffler distributions nests the exponential distribution (Haubold
et al., 2011), and hence the fractional Poisson process generalizes the
standard Poisson process. Again as the threshold size increases and the
threshold crossing probability \(p\) decreases, the fractional Poisson
process is rarefied: The scale parameter of the Mittag-Leffler
inter-arrival times of threshold crossing times increases, but
\emph{superlinearly}; see the Theorem below.

Maxima of events which occur according to a renewal process with
heavy-tailed waiting times have been studied under the names
``Continuous Time Random Maxima process'' (CTRM) (Benson et al., 2007;
Hees and Scheffler, 2018a, 2018b; Meerschaert and Stoev, 2008),
``Max-Renewal process'' (Basrak and Špoljarić, 2015; Silvestrov, 2002;
Silvestrov and Teugels, 2004), and ``Shock process'' (Anderson, 1987;
Esary and Marshall, 1973; Gut and Hüsler, 1999; Shanthikumar and Sumita,
1985, 1984, 1983). The existing literature focuses on probabilistic
results surrounding these models. In this work, however, we introduce a
method of inference for this type of model, which is seemingly not
available in the literature.

We review the marked renewal process in Section 2, and derive a scaling
limit theorem for inter-exceedance times in Section 3. We give a
statistical procedure to estimate model parameters via stability plots
in Section 5, but to set the stage we first discuss inference for the
Mittag-Leffler distribution in Section 4. A simulation study of the
effectiveness of our statistical procedure is given in Section 6. In
Section 7 we apply our method to a real data set. In Section 8, we
discuss the memory property of the Mittag-Leffler distribution, and how
it affects the predictive distribution for the time until the next
threshold crossing event. Finally we close with a discussion and
conclusion in Section 9. For all statistical computations we have used R
(R Core Team, 2018). All code and data used for the analysis in this
article has been organized into an R package \texttt{CTRE}
(\url{https://github.com/strakaps/CTRE}). The source code for the
figures generated in this manuscript is available online at
\url{https://github.com/strakaps/bursty-POT}.

@@

In this article, we restrict ourselves to the uncoupled case, where
\(W\) and \(J\) are independent. Then the two sequences
\(X(\ell, n)_{n \in \mathbb N}\) and \(T(\ell, n)_{n \in \mathbb N}\)
are independent as well. To see why, note that \(X(\ell)\) is, in
distribution, simply equal to \(J | J > \ell\), independent of any
waiting time \(W_k\).\\
Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top
panel, where \(W\) has a stable distribution with tail parameter
\(\beta =\) 0.8 (and skewness \(1\) and location \(0\)), and where \(J\)
is from a standard Gumbel distribution. In the bottom panel, we plot a
time series of solar flare intensities derived from a NASA dataset
(Dennis et al., 1991) which we will later examine more closely (see
Section 7). Clearly, the simulated data exhibit long intervals
\emph{without any} events, whereas in the real-world dataset events
appear continuously. The threshold exceedances, however, appear to have
visually similar statistical behaviour in both models. Observations
below a threshold are commonly discarded in Extreme Value Theory (POT
approach); likewise, the CTRE model interprets these observations as
noise and discards them.

@@

In this section we state and prove the key theorem, see below. For an
accessible introduction to regular variation and stable limit theorems,
we recommend the book by Meerschaert and Sikorskii (2012).

@@

\emph{Proof of Theorem:} We interpret the threshold crossing time
\(T(\ell)\) as the hitting time of the underlying CTRM (Continuous Time
Random Maxima) or ``max-renewal process'', and then utilize a result by
Meerschaert and Stoev (2008). The running maximum process is defined as
\[
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
\] and since we assume that the \(J_k\) have a continuous distribution,
there exist norming functions \(a(c)\) and \(d(c)\) such that \[
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
\] where \(F\) is a generalized extreme value distribution, and
\(\ell^*\) is any value from the support of \(F\). The CTRM process is
then defined via \[
V(t) = M(N(t)), \quad t \ge 0
\] where \(N(t)\) is the renewal process associated with the waiting
times \(W_k\): \[
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
\] Now a key observation is that \[
T(\ell) = \inf\{t: V(t) > \ell\}, 
\] and that \[
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
\] By (Theorem 3.1, Meerschaert and Stoev, 2008), we have the stochastic
process convergence \[
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
\] where \(Y(t)\) is a time-changed (``subordinated'') extremal process,
and where \(\tilde b(c)\) is a regularly varying norming function which
is \emph{inverse} to \(b(c)\), in the sense that
\(b(\tilde b(c)) \sim c \sim \tilde b(b(c))\).

Without loss of generality, we choose \(\ell^*\) such that
\(F(\ell^*) = 1/e\), and let
\(\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))\). We may then calculate
\[
\mathbf P\left[ \frac{T(\ell)}{b(1/p)} > t \right]
= \mathbf P[T(\ell) > b(1/p) t]
= \mathbf P[V(ct) \le \ell]
\] where we have substituted \(c = b(1/p)\). Moreover \[
\mathbf P[V(ct) \le \ell]
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
\] Defining the hitting time of level \(\ell^*\) by \(Y(t)\) as
\(\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}\), we then have \[
P\left[ Y(t) \le \ell^* \right] = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
\] by (Proposition 4.2, Meerschaert and Stoev, 2008), where \(X\) is an
exponential random variable with mean \(1\). Using (Theorem 19.1,
Haubold et al., 2011), we see that
\(X^{1/\beta} D \sim {\rm ML}(\beta, 1)\), concluding the proof. \qed

For a scale parameter \(\sigma > 0\), we write
\({\rm ML}(\beta, \sigma)\) for the distribution of \(\sigma W_\beta\).
The Mittag-Leffler distribution with parameter \(\beta \in (0,1]\) is a
heavy-tailed positive distribution for \(\beta < 1\), with infinite
mean. However, as \(\beta \uparrow 1\), \({\rm ML}(\beta, \sigma)\)
converges weakly to the exponential distribution \({\rm Exp}(\sigma)\)
with mean \(\sigma\). This means that although its moments are all
infinite, the Mittag-Leffler distribution may (if \(\beta\) is close to
1) be indistinguishable from the exponential distribution, for the
purposes of applied statistics. For a detailed reference on the
Mittag-Leffler distribution, see e.g.~ Haubold et al. (2011), and for
algorithms, see e.g.~the R package \texttt{MittagLeffleR} (Gill and
Straka, 2017).

@@

Historically, the first method proposed for the estimation of the
Mittag-Leffler distribution parameters was the fractional moment
estimator by Kozubowski (2001). Unlike the first moments, the fractional
moments of order \(p\) for \(p<\beta\) exist and are tractable. One
drawback of this method is that constant priors for the tail parameter
are needed for the calculation of the estimates. Cahoy et al. (2010)
proposed a moment estimator of the log-transformed data, which does not
require any prior. Furthermore, they performed simulation studies
illustrating that the log-moment outperforms the fractional moment
estimator with respect to bias and root mean squared error (RMSE).

Due to the form of the density function of a Mittag-Leffler distribution
there exists no closed form for the Maximum Likelihood estimator (MLE).
In the R Package \texttt{MittagLeffleR} (Gill and Straka, 2017), Maximum
Likelihood estimation is implemented via numerical optimization. The MLE
slightly outperforms the log-moment estimator regarding bias and RMSE
for big enough sample sizes, but is extremely computational intensive.
Figure \ref{fig:MSE} shows that both estimators perform well, even for
small sample sizes.

@@

Since the Mittag-Leffler distribution is heavy-tailed, many researchers
would intuitively give the highest importance to the tail behaviour of
the distribution. Of course, one can also use established tail exponent
estimators for the estimation of the parameter \(\beta\) such as the
Hill estimator (Hill, 1975) based on the \(r+1\) upper order statistics
\begin{align}\label{eq:hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align} where \(X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}\)
denote the order statistics in decreasing order of a sequence
\(X_1,\ldots,X_n\). However, these methods are statistically less
efficient since they only use a portion of the information contained in
the data, as mentioned by Kozubowski (2001). Moreover, the hill
estimator requires a tuning parameter, denoted with \(r\) in
\eqref{eq:hill}. Additionaly, the estimator only performs reliably well
for distributions close to Pareto (see Resnick, 1997). Figure
\ref{fig:Hillplots} shows Hill plots for Mittag-Leffler simulated data,
with varying sample sizes and tails. To deduce the correct tail
parameter estimates from these plots is virtually impossible. To
estimate the tail parameter of a Mittag-Leffler distribution with the
hill estimator, the sample size has to be much larger and one has to
choose a large tuning parameter \(r\). In case that only events above a
high threshold are recorded, even \(n=1000\) events are unrealistic.
Furthermore the Hill estimator of course completely fails if the
inter-arrival times are exponential distributed and not heavy-tailed. We
will comeback to the problems with the hill estimtator in Section
\ref{Simulationstudy}.

@@

Since the exponential distribution is nested in the Mittag-Leffler
family of distributions, a Likelihood-ratio Test (LRT) seems to be an
appropriate way to choose between a model with exponential and
Mittag-Leffler inter-exceedance times. Although the two models are
nested, the asymptotic distribution is not \(\chi^2_1\)-distributed, and
Wilk's Theorem does not hold: under \(H_0\), the parameter \(\beta\) of
Mittag-Leffler distribution is equal to \(1\), and hence lies on the
boundary of the parameter space \((0,1]\). Instead, a valid approach is
a bootstrapped Likelihood-ratio test (see e.g.~Davison et al., 1997).
Figure \ref{Fig:LRT} displays the (simulated) power for the bootstrapped
LRT for Mittag-Leffler distributions with varying tail parameters based
on 1000 simulation runs. As expected, the power decreases for tail
parameters close to one, since the Mittag-Leffler distribution converges
as \(\beta \uparrow 1\) to an exponential distribution; it becomes hard
to differentiate a Mittag-Lefller distribution from an exponential.

@@

The estimates \(\hat \beta\) and \(\hat \sigma_0\) give an estimate of
the distribution of exceedance times, dependent on the threshold
\(\ell\): \begin{align*}
T(\ell) \sim {\rm ML}(\hat \beta, k^{-1/\hat \beta} \hat \sigma_0),
\end{align*} where \(\ell\) equals the \(k\)-th order statistic. For
quick estimates of the Mittag-Leffler parameters we have used the method
of log-transformed moments (Cahoy, 2013). We have verified the validity
of our estimation algorithm via simulations, see Section
\ref{Simulationstudy}.

@@

Figure \ref{fig:flares} shows the stability plots for the solar flare
data, on the left for the tail parameter and on the right for the scale
parameter. Dotted lines show 95\% confidence intervals, which are
derived from the asymptotic normality of the log-moments estimators
(Cahoy, 2013) and the \(\delta\)-method (Gill and Straka, 2017), dashed
lines show the deduced values of \(\beta\) resp.~\(\sigma_0\). The
stability plot for the tail stabilizes nicely round about 0.85 (dashed
line), while it is a little bit harder to deduce an estimate for the
scale parameter. Howerever, the stability plot for the scale parameter
seems to stabilize round about \(3 \times 10^7\) (dashed line).

@@

The bursty CTRE approach to model ``non-Poissonian'' threshold crossing
times should be contrasted with the (now standard) approach of clusters
of extremes, see e.g.~Ferro and Segers (2003). In this approach, i.i.d.
event sequences of magnitudes are generalized to stationary sequences of
event magnitudes (subject to a mixing condition). The two approaches are
fundamentally different: A clustering model assumes that each event
belongs to one particular (latent) group of events. For bursts, however,
the aim is to identify an underlying scale-free pattern in the event
dynamics, which is often characteristic of complex systems. It is an
interesting open problem to develop quality criteria, based e.g.~on
measures of surprise (Lee et al., 2015), which guide an applied
statistician in the choice between a clustering and a CTRE approach for
a particular problem. Moreover, we believe it may be possible to unify
the two approaches by considering CTREs based on MRPs with a
\emph{stationary}, rather than i.i.d., sequence of magnitudes.

Finally, a purely scale-free pattern for event times may be too rigid as
an assumption for some bursty time series, because often the
heavy-tailed character of the inter-arrival time distribution does not
hold at all time scales; rather, it applies at short and intermediate
time scales, and is truncated (or tempered, reverting to an exponential
distribution) at very long time scales (see e.g.~Meerschaert et al.,
2012; and Aban et al., 2006). In such situations, a ``tempered''
Mittag-Leffler distribution may provide a more realistic fit, which we
aim to introduce in follow-up work.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Aban06}{}%
Aban, I.B., Meerschaert, M.M., Panorska, A.K., 2006. Parameter
estimation for the truncated pareto distribution. J. Am. Stat. Assoc.
101, 270--277.
doi:\href{https://doi.org/10.1198/016214505000000411}{10.1198/016214505000000411}

\leavevmode\hypertarget{ref-Anderson1987}{}%
Anderson, K.K., 1987. Limit Theorems for General Shock Models with
Infinite Mean Intershock Times. J. Appl. Probab. 24, 449--456.

\leavevmode\hypertarget{ref-Bagrow2013}{}%
Bagrow, J.P., Brockmann, D., 2013. Natural emergence of clusters and
bursts in network evolution. Phys. Rev. X 3, 1--6.
doi:\href{https://doi.org/10.1103/PhysRevX.3.021016}{10.1103/PhysRevX.3.021016}

\leavevmode\hypertarget{ref-Barabasi2005}{}%
Barabási, A.L., 2005. The origin of bursts and heavy tails in human
dynamics. Nature 435, 207--211.
doi:\href{https://doi.org/10.1038/nature03459}{10.1038/nature03459}

\leavevmode\hypertarget{ref-Basrak2014}{}%
Basrak, B., Špoljarić, D., 2015. Extremes of random variables observed
in renewal times. Stat. Probab. Lett. 97, 216--221.
doi:\href{https://doi.org/10.1016/j.spl.2014.11.025}{10.1016/j.spl.2014.11.025}
